---
title: "NanoString Analysis"
author: "benjamin.tk.yuen"
date: "04/08/2025"
output: html_document
---
Analysis of NanoString data. In this example, I look at data from multiple patients, patient cohorts, and visits.

#Packages
```{r load_packages, include = FALSE}
# Load required packages
packages_to_load <- c("dplyr", "stringr", "readr", "pheatmap", "ggplot2", 
                      "tidyr", "data.table", "rtracklayer", "xCell", 
                      "edgeR", "limma", "NanoStringNorm", "SummarizedExperiment",
                      "ggrepel", "msigdbr", "msigdbdf", "clusterProfiler",
                      "enrichplot", "randomcoloR", "purrr", "tibble", 
                      "factoextra")
lapply(packages_to_load, FUN = require, character.only = TRUE)
```

#Gene Signatures
```{r load_gene_sig, include = FALSE}

# Define gene signatures from literature
t_cell_cd8 <- c("CD8A", "CD8B", "GZMA", "GZMB", "GZMK", "PRF1", "EOMES", "LAG3", "IFNG",
                "CD3D", "CD3E", "CD3G", "CD247", "CD2", "CD5", "CD7",
                     "CD4", "CD8A", "CD8B", "CD28", "LCK", "ZAP70", "IL2RA") %>% unique()

nk_cell <- c("KLRD1", "KLRC1", "KLRC2", "KLRF1", "NKG7", "GNLY", "FCGR3A", "GZMA", "GZMB", 
             "KLRD1", "KLRC1", "KLRC2", "KLRK1", "NCR1", "NCR3", 
                       "FCGR3A", "GNLY", "GZMA", "GZMB", "NKG7") %>% unique()

cytotoxic_cell <- c("GZMA", "GZMB", "GZMH", "GZMK", "PRF1", "GNLY", "NKG7", "KLRD1", "FASL", "TNFSF10") %>% unique()

t_cell <- c("CD3D", "CD3E", "CD3G", "CD247", "CD2", "CD5", "IL7R", "CD28") %>% unique()

dendritic_cell <- c("ITGAX", "HLA-DRA", "HLA-DPA1", "HLA-DPB1", "CD1C", "CD83", "CD86", "FLT3", "CLEC9A") %>% unique()

b_cell <- c("CD19", "CD79A", "CD79B", "MS4A1", "CD22", "BLK", "BLNK") %>% unique()

th1_cell <- c("TBX21", "IFNG", "IL12RB2", "CXCR3", "STAT1", "STAT4") %>% unique()

th2_cell <- c("GATA3", "IL4", "IL5", "IL13", "CCR4", "STAT6") %>% unique()

t1_ifn <- c("BST2", "CCL2", "CXCL10", "IFI35", "IFIT1", "IFIT2", "IFITM1", "IRF7", "ISG15", "ISG20", "LAMP3", "MX1",
            "OAS3", "STAT1", "STAT2", "IFNA2") %>% unique()

t_cell_activated <- c("CD8A", "CD3E", "CD3G", "GZMA", "PRF1", "IL7R", "SLAMF6", "ICOS", "CD40LG", "TNFSF8", "CXCR5",
                      "TNFSF12", "TNFRSF14", "TBX21", "STAT4", "CXCR3") %>% unique()

ifng_sig <- c("IFNG", "CCL5", "CD27", "CMKLR1", "CXCL9", "CXCR6", "HLA-E", "IDO1", "PSMB10", "STAT1") %>% unique()

antigen_processing <- c("HLA-A", "HLA-B", "HLA-C", "B2M", "TAP1", "TAP2", "ERAP1",
                        "HLA-DRA", "HLA-DRB1", "HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DQB1",
                        "CD74", "CALR", "CANX", "PDIA3")

macrophage_functions <- c("CD14", "CD68", "CD163", "MRC1", "CSF1", "CSF1R", 
                          "TNF", "IL1B", "IL6", "ARG1", "MARCO", "CD36")

pathogen_defense <- c("IL1B", "TNF", "IL6", "IFNG", "CAMP", "DEFB1", "NOS2", "NOD2", "MYD88")

tlr_pathway <- c("TLR1", "TLR2", "TLR3", "TLR4", "TLR5", "TLR6", "TLR7", 
                 "TLR8", "TLR9", "TLR10", "MYD88", "TIRAP", "TRIF", "TRAM")

cytotoxic_genes <- c("GNLY", "GZMB", "PRF1", "GZMA", "GZMH", "GZMM", "GZMK")

antigen_processing_modified <- c("THBS1", "PSMB9", "HLA-A", "CD8A", "TAP1", "HLA-C", "HLA-B", "CD1E", "PSMB7", "TAP2", "MR1", "TAPBP")

signature_list <- list("CD8 T-cells" = t_cell_cd8,
                       "NK cells" = nk_cell,
                       "Cytotoxic cells" = cytotoxic_cell,
                       "T-cells" = t_cell,
                       "DCs" = dendritic_cell,
                       "B-cells" = b_cell,
                       "Th1 cells" = th1_cell,
                       "Th2 cells" = th2_cell,
                       "Type 1 IFN cells" = t1_ifn,
                       "Activated T-cells" = t_cell_activated,
                       "IFNg Signature cells" = ifng_sig,
                       "Antigen Processing" = antigen_processing,
                       "Macrophage Signature" = macrophage_functions,
                       "TLR Pathway" = tlr_pathway,
                       "Pathogen Defense" = pathogen_defense,
                       "Cytotoxic Molecules" = cytotoxic_genes,
                       "Class I Antigen Processing" = antigen_processing_modified)

signature_list_2 <- list("Th1 cells" = th1_cell, 
                        "Th2 cells" = th2_cell,
                        "Type 1 IFN cells" = t1_ifn, 
                        "Activated T-cells" = t_cell_activated,
                        "IFNg Signature cells" = ifng_sig)

# Load the NanoString CancerImmune probes
# The first line of the .csv is a descriptor and can be skipped 
ns_probe_annotations <- read.csv("/home/byuen/references/nanostring/ProbeAnnotations_NS_CancerImmune_V1.1.csv", skip = 1)

ns_probe_list <- strsplit(ns_probe_annotations$Related.Probes, ";") %>% unlist() %>% unique()
```

#Functions
```{r load_functions, include = FALSE}

# Function to calculate geometric mean from a row of data
calc_geom_mean <- function(data_row, na.rm = TRUE) {
  
  # Filter positive values only (and if none remain, return NA)
  x_pos <- data_row[data_row > 0]
  
  if(length(x_pos) == 0) 
    return(NA_real_)
  
  exp(mean(log(x_pos), na.rm = na.rm))
}

# Function to calculate a Z-score from a row of data
calc_group_z_score <- function(data_row){
  (data_row - mean(data_row)) / sd(data_row)
}

# Function to generate a heatmap relative to baseline sample
make_gene_heatmap <- function(probeset_labels, probeset, baseline_visit, plot_data, file_id, output_path){
    
    # Customer wants log2 change around baseline sample in subsequent visits
    # Use arithmetic mean log2(visit) - arithmetic mean log2(baseline) 
    #   to determine the change
    # See NanoString manual 'MAN-C0011-04 Gene Expression Data Analysis 
    #   Guidelines' for more details (p.19)
    
    # Set up baseline vs. treatment groups, add baseline identifiers to gene names in baseline group
    temp_baseline <- plot_data %>% filter(visit_id == baseline_visit) %>% 
                                    select(patient_id, visit_id, id, order, cohort, all_of(probeset_labels)) %>%
                                    mutate(across(all_of(probeset_labels), as.numeric)) %>%
                                    rename_with(~ paste0(., "_baseline"), all_of(probeset_labels)) %>%
                                    mutate(treatment = "baseline")
    
    temp_treatment <- plot_data %>% filter(visit_id != baseline_visit) %>% 
                                    select(patient_id, visit_id, id, order, cohort, all_of(probeset_labels)) %>%
                                    mutate(across(all_of(probeset_labels), as.numeric)) %>%
                                    mutate(treatment = "treated")

    # Generate new data table based on difference between gene on treatment timepoint and baseline
    temp_data <- temp_treatment %>% left_join(temp_baseline, by = "patient_id") %>%
      mutate(across(all_of(probeset_labels), 
                    .fns = ~ . - get(paste0(cur_column(), "_baseline")),
                    .names = "log2diff_{.col}"))
                      
    # Remove extra columns (baseline and treated values)
    temp_data <- temp_data %>% select(paste0("log2diff_", probeset_labels), 
                                              patient_id, visit_id.x, id.x, 
                                              order.x, cohort.x) %>% 
                                arrange(order.x) %>% as.data.frame()
    
    # Remove new column identifiers
    colnames(temp_data) <- gsub("log2diff_", "", colnames(temp_data))
    colnames(temp_data) <- gsub(".x", "", colnames(temp_data))
    rownames(temp_data) <- temp_data$id
    
    # Generate heatmaps based on predefined order
    px_list <- temp_data %>% pull(patient_id)
    px_gaps <- which(px_list[-1] != px_list[-length(px_list)])
    
    # Determine the maximum absolute value for scaling (maximum of pos or neg directions)
    max_val <- max(abs(temp_data %>% select(!c(patient_id, visit_id, id, order, cohort)) %>% as.matrix()), na.rm = TRUE)
    
    color_map <- colorRampPalette(c("darkblue", "white", "darkred"))(100)
    
    color_break <- seq(-max_val, max_val, length.out = length(color_map) + 1)
   
    plot_heatmap_order <- pheatmap(temp_data %>% select(!c(patient_id, visit_id, id, order, cohort)) %>% as.matrix(),
                            annotation_row = temp_data %>% arrange(order) 
                                %>% select(cohort, visit_id) %>% mutate(cohort = as.factor(cohort)),
                            cluster_rows = FALSE, # Preserve patient order
                            show_rownames = TRUE,
                            show_colnames = TRUE,
                            gaps_row = px_gaps,
                            scale = "none",
                            color = color_map,
                            breaks = color_break,
                            main = paste0("Log2FC from baseline for ", probeset, " Signature"))
    
    # Generate heatmaps based on patient visit
    px_list <- temp_data %>% arrange(visit_id) %>% pull(visit_id)
    px_gaps <- which(px_list[-1] != px_list[-length(px_list)])
    
    plot_heatmap_visit <- pheatmap(temp_data %>% arrange(visit_id) %>% 
                                     select(!c(patient_id, visit_id, id, order, cohort)) %>% as.matrix(),
                            annotation_row = temp_data %>% arrange(order) 
                                %>% select(cohort, visit_id) %>% mutate(cohort = as.factor(cohort)),
                            cluster_rows = FALSE, # Preserve patient order
                            show_rownames = TRUE,
                            show_colnames = TRUE,
                            gaps_row = px_gaps,
                            scale = "none",
                            color = color_map,
                            breaks = color_break,
                            main = paste0("Log2FC from baseline for ", probeset, " Signature"))
    
    if(!dir.exists(paste0(file.path(output_path), "heatmap_by_order"))){
      dir.create(paste0(file.path(output_path), "heatmap_by_order"), recursive = TRUE)
    }
    
    if(!dir.exists(paste0(file.path(output_path), "heatmap_by_visit"))){
      dir.create(paste0(file.path(output_path), "heatmap_by_visit"), recursive = TRUE)
    }
  
    png(filename = paste0(file.path(output_path), "heatmap_by_order/heatmap_log2fc_", file_id,"_", probeset, ".png"), 
        width = (800 + length(probeset_labels)*60), height = (800 + length(temp_data %>% select(id) %>% length())*60), res = 150)
    grid::grid.newpage()
    grid::grid.draw(plot_heatmap_order$gtable)
    dev.off()
    
    png(filename = paste0(file.path(output_path), "heatmap_by_visit/heatmap_log2fc_", file_id,"_", probeset, ".png"), 
        width = (800 + length(probeset_labels)*60), height = (800 + length(temp_data %>% select(id) %>% length())*60), res = 150)
    grid::grid.newpage()
    grid::grid.draw(plot_heatmap_visit$gtable)
    dev.off()
    
    return(temp_data)
}

# Function to make a gene dotplot
make_gene_dotplot <- function(plot_data, probeset, probeset_labels, file_id){
  
  # Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
  plot_data$geo_mean <- apply(plot_data %>%
                          mutate(across(all_of(probeset_labels), as.numeric)) %>%
                          select(any_of(probeset_labels)) %>%
                          as.matrix(),
                          1, calc_geom_mean)
  
  plot_order <- plot_data %>% arrange(timepoint) %>% pull(visit_id) %>% unique()
  
  plot_data <- plot_data %>%
    mutate(visit_id = factor(visit_id, levels = plot_order)) %>% 
    mutate(cohort = as.numeric(cohort))
  
  plot_ids <- unique(plot_data$id)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)
  
  plot_cohorts <- sort(unique(plot_data$cohort))
  color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
  color_map <- setNames(color_values, plot_cohorts)

  # Plot
  plotted_data <- ggplot(plot_data, aes(x = visit_id, y = geo_mean)) +
    geom_line(aes(group = id, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = id, color = factor(cohort), shape = id), size = 4, stroke = 1.2, fill = "white") +
    theme_bw() +
    scale_shape_manual(name = "id", values = shape_map) + 
    scale_color_manual(name = "cohort", values = color_map) +
    labs(title = paste0(probeset, " signature over visit_id by patient_id"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    ylim(min(plot_data$geo_mean) - 1, max(plot_data$geo_mean) + 1) +
    theme(legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1))

  plot(plotted_data)
  
  if(!dir.exists(paste0(working_dir, "dotplot_by_visit"))){
    dir.create(paste0(working_dir, "dotplot_by_visit"), recursive = TRUE)
  }
  
  png(filename = paste0(working_dir, "dotplot_by_visit/dotplot_", file_id, "_", probeset, ".png"))
  plot(plotted_data)
  dev.off()
}

# Function to generate a heatplot using enrichplot
make_heatplot <- function(gene_set, degs, title){
  
  top_terms = 5
  
  ordered_terms <- order(gene_set@result$Count, decreasing = TRUE)
  idx <- ordered_terms[seq_len(min(top_terms, length(ordered_terms)))]
  
  res_df <- gene_set
  res_df@result <- gene_set@result[idx, , drop = FALSE]
  term_ids <- res_df@result$ID
  res_df@geneSets  <- res_df@geneSets[term_ids]
  
  gene_list <- degs$logFC %>% as.numeric()
  names(gene_list) <- rownames(degs)
  
  top_desc <- res_df@result %>%
    arrange(desc(Count)) %>%
    pull(Description) %>%
    head(top_terms)
  
  res_df@result$Description <- factor(res_df@result$Description,
                                  levels = top_desc)

  plotted_data <- heatplot(
    res_df,
    foldChange = gene_list,
    showCategory = top_terms) + 
    ggtitle(title) +
    scale_fill_gradient2(
            low = "darkblue",
            mid = "white",
            high = "darkred",
            midpoint = 0,
            name = expression(log[2]~FC)) +
    scale_y_discrete(limits = rev(term_ids))
  
  return(plotted_data)
}

# Function to save the ORA information 
save_ora_data <- function(overrep_data, de_genes, group_var, split_var, group2, group1, split, output_path){
  
  if(!is.null(overrep_data)){
    if(!overrep_data@result %>% nrow() == 0){
      write.csv(x = overrep_data %>% as.data.frame() %>% head(10),
        file = paste0(file.path(output_path), "gsea/msig_c7_overrepresented_genes_", group_var, "_", 
          group2, "_vs_", group1, "_@_", split_var, "_", split, ".csv"))
                  
        temp_plot <- cowplot::plot_grid(
          enrichplot::dotplot(overrep_data) + 
          ggtitle(paste0("Over-represented genes in ", group_var, " ", 
            group2, "\nfound in ImmuneSigDB C7 categories\nfor ", split_var, "_", split)) +
          scale_size_continuous(breaks = seq(0, max(overrep_data@result$Count, na.rm = TRUE), by = 1)))
                  
        plot(temp_plot)
                
        png(filename = paste0(file.path(output_path), "gsea/msig_c7_overrepresented_genes_", group_var, "_",
          group2, "_vs_", group1, "_@_", split_var, "_", split, ".png"), 
          width = 1000, height = 1000, res = 150)
        plot(temp_plot)
        dev.off()
                  
        temp_plot <- make_heatplot(gene_set = overrep_data, degs = de_genes,
          title = paste0("Over-represented genes in ", group_var, " ", 
            group2, "\nfound in ImmuneSigDB C7 categories\nfor ", split_var, "_", split))
        
        png(filename = paste0(file.path(output_path), "gsea/msig_c7_overrepresented_genes_heatmap_", group_var, "_",
          group2, "_vs_", group1, "_@_", split_var, "_", split, ".png"), 
          width = 2000, height = 1000, res = 150)
        plot(temp_plot)
        dev.off()
    }
  }
}

# Function bringing together DEGs, ORA, and GSEA
#` Main wrapper
#' @expr_mat is a gene matrix with samples as rows and genes as cols
#' @meta is metadata, with @sample_name as rownames and containing
#'    @group_var , @split_var , and @sample_name as cols
#' @split_var is the initial group we're comparing (e.g. 'visit_id')
#' @group_var is the secondary group we're comparing within (e.g. 'cohort')
run_group_by_split_DE <- function(expr_mat, meta,
                                  group_var, split_var, output_path) {
  
  # Initialize data frame to return, this will contain all the groups and comparisons
  de_genes_to_return <- data.frame()
  
  # 1) Align meta and expression
  
  # Determine all metadata for group_var and split_var
  meta2 <- meta %>%
    filter(sample_name %in% rownames(expr_mat)) %>%
    arrange(!!sym(split_var), !!sym(group_var))
  
  # Reformat expression matrix so that genes are rows, samples are cols
  expr2 <- expr_mat[meta2$sample_name, , drop = FALSE]
  
  #' 2) Loop through first level (e.g. @split_var or 'timepoint')
  
  # Identify the levels to cycle through
  split_levels <- meta2 %>% 
                    distinct(!!sym(split_var)) %>%
                    pull(!!sym(split_var)) %>% 
                    set_names()
  
  for(sp in split_levels) {
    
      print(paste0("Running level ", sp))
    
      #' @sp is subset of comparison or one 'timepoint'
      idx    <- which(meta2[[split_var]] == sp) #' Location of @sp or 'timepoint' 
      meta_s <- meta2[idx, ] # Subset of metadata
      expr_s <- expr2[idx, , drop = FALSE] # Subset of expression matrix
      
      #' ID groups present within @sp or 'timepoint'
      grps <- unique(as.character(meta_s[[group_var]]))
      
      # Only proceed if there are >2 groups to compare
      if (length(grps) > 1) {
        
        # Build all pairs
        pairs <- combn(grps, 2, simplify = FALSE)
        
        #' 3) Run each pair through differential expression and add onto @de_genes_to_return

        for(pair in pairs){
          
          #if(length(pairs) == 1){
          #  pair <- pairs[[1]]
          #}
          
          # Subset metadata on group or 'cohort' for the 'timepoint'
          meta_sub <- meta_s %>% filter(!!sym(group_var) %in% unlist(pair)) 
          
          # Subset expression matrix on group or 'cohort' for the 'timepoint'
          expr_sub <- expr_s[meta_sub$sample_name, , drop = FALSE] %>% mutate(across(everything(), as.numeric))
          
          # ID groups and only proceed if greater than 2 present
          grp <- factor(paste(group_var, meta_sub[[group_var]], sep = "_"))
          lvl <- levels(grp)
          
          if (length(lvl) > 1) {
            
              print(paste0("More than 2 points per group, running ", lvl[1], " with ", lvl[2]))  
              
              design <- model.matrix(~0 + grp)
              colnames(design) <- lvl
              
              fit <- limma::lmFit(t(expr_sub), design)
              
              contrast_id <- paste0(lvl[2], " - ", lvl[1]) # Descriptor for comparison
              
              # Initialize the contrasts vs. levels 
              cont <- limma::makeContrasts(
                  contrasts = contrast_id,
                  levels = design
              )
              
              fit2 <- limma::eBayes(contrasts.fit(fit, cont))
              
              de_genes <- limma::topTable(fit2, 
                coef = contrast_id, 
                number = Inf, 
                sort.by = "P",
                adjust.method = "BH") 
              
              de_genes_to_return <- rbind(
                de_genes_to_return, 
                de_genes %>%
                  as_tibble(rownames = "gene") %>%
                  mutate(group1 = lvl[1],
                    group2 = lvl[2],
                    comparison = paste0(lvl[2], "_vs_", lvl[1], "_@_", sp))
                    )
                
              
              if(!dir.exists(paste0(file.path(output_path), "volcanoplot"))){
                dir.create(paste0(file.path(output_path), "volcanoplot"), recursive = TRUE)
              }
              
              write.csv(x = de_genes,
                file = paste0(working_dir, "volcanoplot/de_genes_", group_var, "_",
                lvl[2], "_vs_", lvl[1], "_@_", split_var, "_", sp, ".csv"))
              
              # Create data for volcano plot
              volcano_data <- de_genes %>%
                as_tibble(rownames = "gene") %>%
                mutate(
                  negLogP = -log10(P.Value),
                  sig = case_when(
                          logFC >=  1 & P.Value < 0.05 ~ "Up",
                          logFC <= -1 & P.Value < 0.05 ~ "Down",
                          TRUE ~ "Not Selected"
                          )
                      )
              
              temp_plot <- plot_volcano(volcano_data = volcano_data,
                                        title = paste0(
                                          "Genes Up/Down-regulated in\n", group_var, " ",
                                          lvl[2], " (group 2) vs. ", lvl[1], " (group 1)\n",
                                          " for ", split_var, "_", sp))
              plot(temp_plot)
            
              png(filename = paste0(file.path(output_path), "volcanoplot/volcano_log2_", group_var, "_",
                                    lvl[2], "_vs_", lvl[1], "_", split_var, "_", sp, ".png"), 
                      width = 1000, height = 1000, res = 150)
              plot(temp_plot)
              dev.off()

              group2_degs <- de_genes %>% filter(logFC >= 1 & P.Value < 0.05) %>% rownames()
              group1_degs <- de_genes %>% filter(logFC <= -1 & P.Value < 0.05) %>% rownames()

              msigc7_data <- msigdbr::msigdbr(species = "Homo sapiens", collection = "C7") %>% select(gs_name, gene_symbol)
              
              overrep_data_group2 <- clusterProfiler::enricher(
                gene = group2_degs,
                TERM2GENE = msigc7_data,
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.25)
            
              overrep_data_group1 <- clusterProfiler::enricher(
                  gene = group1_degs,
                  TERM2GENE = msigc7_data,
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.25)
              
              if(!dir.exists(paste0(file.path(output_path), "gsea"))){
                dir.create(paste0(file.path(output_path), "gsea"), recursive = TRUE)
              }
              
              # Save the ORA information
              save_ora_data(overrep_data = overrep_data_group2, de_genes = de_genes,
                            group_var = group_var, group2 = lvl[2], group1 = lvl[1],
                            split_var = split_var, split = sp,
                            output_path = output_path)
              
              save_ora_data(overrep_data = overrep_data_group1, de_genes = de_genes,
                            group_var = group_var, group2 = lvl[1], group1 = lvl[2],
                            split_var = split_var, split = sp,
                            output_path = output_path)
              
          } # End of if statement for group_var > 1
          
        } # End of loop for group_var
        
      } # End of if statement for split_var groups > 1
      
    } # End loop for split_var
  
  write.csv(x = de_genes_to_return,
            file = paste0(file.path(output_path), "de_genes_all_", split_var, "_by_", group_var, ".csv"))
  
  return(de_genes_to_return)
}

# Quick and dirty way to plot individual cell profiler scores
plot_celltype_score  <- function (cellprofiler_input, celltype, data_type, timepoint_levels, output_path){
  
  celltype_score <- cellprofiler_input %>%
    pivot_longer(
      cols = -c(id, cohort, px, tp),
      names_to  = "cell_type",
      values_to = "abundance"
    ) %>%
    mutate(tp = factor(tp, levels = timepoint_levels))
  
  celltype_score$abundance <- as.numeric(celltype_score$abundance)
  
  plot_ids <- unique(celltype_score$px)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)

  temp_plot <- ggplot(celltype_score %>% filter(cell_type == celltype), aes(x = tp, y = abundance)) +
    geom_line(aes(group = px, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = px, color = factor(cohort)),
               size = 4, stroke = 1.2, fill = "white") +
    scale_shape_manual(name   = "Sample ID",
                       values = shape_map) +
    scale_color_brewer(palette = "Set1", name = "Cohort") +
    
    theme_minimal() +
    theme(
      axis.text.x  = element_text(angle = 30, hjust = 1),
      legend.title = element_text(face = "bold")
    ) +
    ylim(min(celltype_score %>% filter(cell_type == celltype) %>% select(abundance)) - 1, 
         max(celltype_score %>% filter(cell_type == celltype) %>% select(abundance)) + 1) +
    labs(
      title = paste0(celltype, " Abundance Over Treatment (", data_type, ")"),
      x     = "Timepoint",
      y     = "Estimated Abundance"
    )
  
  plot(temp_plot)
  
  if(!dir.exists(output_path)){
      dir.create(output_path, recursive = TRUE)
  }
  
  png(filename = file.path(output_path, paste0(celltype, "_", data_type, "_cellprofiler_score.png")), 
        width = 1000, height = 1000, res = 150)
  
  plot(temp_plot)
  
  dev.off()
}

# Function to plot the volcano data
plot_volcano <- function(volcano_data, title){
  
  plot_data <- ggplot(volcano_data, aes(x = logFC, y = negLogP, color = sig)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(
    values = c(Up = "red", Down = "blue", `Not Selected` = "grey"),
    breaks = c("Up","Down","Not Selected")
    ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_text_repel(
    data = filter(volcano_data, sig != "Not Selected"),
    aes(label = gene),
    size = 3,
    box.padding = 0.3,
    max.overlaps = 20
    ) +
  labs(
    title = title,
    x = expression(log[2]~Fold~Change),
    y = expression(-log[10]~P~Value),
    color = "Significance"
  ) +
  theme_minimal()
  
  return(plot_data)
}

#' Function to run PCA on an expression matrix
#' @exp_matrix is an expression matrix, must be samples (rows) x genes (columns)
#' @meta_data is a metadata table, make sure that @sample_name matches rownames of @exp_matrix
#' Will have an initial check to ensure rownames/sample_name are the same and in order
run_pca <- function(exp_matrix, meta_data, output_path){
  
  stopifnot(all(meta_data$sample_name == rownames(exp_matrix))) # Ensure that sample_name and rownames are the same and in order
  
  exp_matrix <- exp_matrix %>% mutate(across(everything(), as.numeric)) # Ensure that data is numeric
  
  pca_results <- prcomp(exp_matrix, 
                  center = TRUE, # Subtract gene‐wise mean so variances are centered around mean
                  scale. = TRUE) # Divide by gene‐wise SD, prevent high variation genes from dominating PCs

  # Build a plotting data.frame
  pca_df <- as_tibble(pca_results$x, rownames = "sample_name") %>%
    left_join(meta_data, by = "sample_name")
  
  if(!dir.exists(output_path)){
      dir.create(output_path, recursive = TRUE)
  }
  
  # Plot PCA by batch
  temp_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(batch), shape = factor(batch))) +
    geom_point(size = 4, stroke = 2, alpha = 0.8) +
    scale_shape_manual(name  = "Batch",
                     values = c(0:20)[1:length(unique(pca_df$batch))]) +
    labs(
      title    = "PCA: Samples Colored by Batch",
      x        = paste0("PC1 (", pct1, "%)"),
      y        = paste0("PC2 (", pct2, "%)"),
      color    = "Batch"
    ) +
    theme_bw()
  
  plot(temp_plot)
  
  png(filename = file.path(output_path, "pca_batch.png"), 
        width = 950, height = 900, res = 150)
      plot(temp_plot)
  dev.off()
  
  # Plot PCA by cohort
  temp_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(cohort), shape = factor(cohort))) +
    geom_point(size = 4, stroke = 2, alpha = 0.8) +
    scale_shape_manual(name  = "Cohort",
                     values = c(0:20)[1:length(unique(pca_df$cohort))]) +
    labs(
      title    = "PCA: Samples Colored by Cohort",
      x        = paste0("PC1 (", pct1, "%)"),
      y        = paste0("PC2 (", pct2, "%)"),
      color    = "Cohort"
    ) +
    theme_bw()
  
  plot(temp_plot)
  
  png(filename = file.path(output_path, "pca_cohort.png"), 
        width = 950, height = 900, res = 150)
      plot(temp_plot)
  dev.off()
  
  # Plot PCA by visit
  temp_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(visit_id), shape = factor(visit_id))) +
    geom_point(size = 4, stroke = 2, alpha = 0.8) +
    scale_shape_manual(name  = "Visit",
                     values = c(0:20)[1:length(unique(pca_df$visit_id))]) +
    labs(
      title    = "PCA: Samples Colored by Visit",
      x        = paste0("PC1 (", pct1, "%)"),
      y        = paste0("PC2 (", pct2, "%)"),
      color    = "Visit"
    ) +
    theme_bw()
  
  plot(temp_plot)
  
  png(filename = file.path(output_path, "pca_visit_id.png"), 
         width = 950, height = 900, res = 150)
      plot(temp_plot)
  dev.off()
  
  # Plot PCA by patient
  temp_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(patient_id), shape = factor(patient_id))) +
    geom_point(size = 4, stroke = 2, alpha = 0.8) +
    scale_shape_manual(name  = "Patient",
                     values = c(0:20)[1:length(unique(pca_df$patient_id))]) +
    labs(
      title    = "PCA: Samples Colored by Patient",
      x        = paste0("PC1 (", pct1, "%)"),
      y        = paste0("PC2 (", pct2, "%)"),
      color    = "Patient"
    ) +
    theme_bw()
  
  plot(temp_plot)
  
  png(filename = file.path(output_path, "pca_patient_id.png"), 
         width = 950, height = 900, res = 150)
      plot(temp_plot)
  dev.off()
  
  # Plot PCA by with visit, but patient distinguished by shape
  temp_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(visit_id), shape = factor(patient_id))) +
    geom_point(size = 4, stroke = 2, alpha = 0.8) +
    scale_shape_manual(name  = "Patient",
                     values = c(0:20)[1:length(unique(pca_df$patient_id))]) +
    labs(
      title    = "PCA: Samples Colored by Visit",
      x        = paste0("PC1 (", pct1, "%)"),
      y        = paste0("PC2 (", pct2, "%)"),
      color    = "Visit"
    ) +
    theme_bw()
  
  plot(temp_plot)
  
  png(filename = file.path(output_path, "pca_visit_id_by_patient_id.png"), 
         width = 950, height = 900, res = 150)
      plot(temp_plot)
  dev.off()
 
  return(pca_results) 
}

make_gene_custom_dotplot <- function(plot_data, probeset, probeset_labels, file_id, shape_map, color_map){
  
  # Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
  plot_data$geo_mean <- apply(plot_data %>%
                          mutate(across(all_of(probeset_labels), as.numeric)) %>%
                          select(any_of(probeset_labels)) %>%
                          as.matrix(),
                          1, calc_geom_mean)
  
  plot_order <- plot_data %>% arrange(timepoint) %>% pull(visit_id) %>% unique()
  
  plot_data <- plot_data %>%
    mutate(visit_id = factor(visit_id, levels = plot_order)) %>% 
    mutate(cohort = as.numeric(cohort))
  
  # Plot
  plotted_data <- ggplot(plot_data, aes(x = visit_id, y = geo_mean)) +
    geom_line(aes(group = id, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = id, color = factor(cohort)), size = 4, stroke = 1.2, fill = "white") +
    theme_bw() +
    scale_shape_manual(name = "id", values = shape_map) + 
    scale_color_manual(name = "cohort", values = color_map) +
    labs(title = paste0(probeset, " signature over visit_id by patient_id"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    ylim(min(plot_data$geo_mean) - 1, max(plot_data$geo_mean) + 1) +
    theme(legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1))

  plot(plotted_data)
  
  if(!dir.exists(paste0(working_dir, "dotplot_by_visit"))){
    dir.create(paste0(working_dir, "dotplot_by_visit"), recursive = TRUE)
  }
  
  png(filename = paste0(working_dir, "dotplot_by_visit/dotplot_", file_id, "_", probeset, ".png"))
  plot(plotted_data)
  dev.off()
}
```

#Load Data
```{r load_data, include = TRUE, echo = FALSE}

working_dir <- "/home/byuen/projects/actymthera/projects/ns_cohort_1-3/results/"
raw_data_dir <- "/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/"

# Load patient NanoString data
log2norm_data <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/log2norm_data.csv")
raw_data <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/raw_data.csv")

# Reshape the data so that genes are in columns and samples are in rows
log2norm_data <- log2norm_data[-c(1,2),] # Remove first two rows, no useful data
log2norm_data <- log2norm_data %>% select(-NS.Probe.ID, -Annotation, -Accession.., -Analyte.Type, 
                                          -Class.Name, -X..Samples.above.Threshold) # Remove unused columns
colnames(log2norm_data) <- gsub("X", "", colnames(log2norm_data)) # Some 'X' characters were added to all the sample names, removing them
log2norm_data <- t(log2norm_data) # Transpose data
colnames(log2norm_data) <- log2norm_data[1,] # Rename columns as gene names
log2norm_data <- log2norm_data[-1,] # Remove first row with gene names

log2norm_data <- as.data.frame(log2norm_data, keep.rownames = TRUE)
log2norm_data$sample_name <- rownames(log2norm_data)

# Load patient identifiers and remove empty columns
patient_ids <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/patient_ids_cohorts_1-3.csv")
patient_ids$sample_name <- gsub(" ", ".", patient_ids$sample_name)
patient_ids$sample_name <- gsub("-", ".", patient_ids$sample_name)

# Merge and organize the data
patient_data <- log2norm_data %>% dplyr::inner_join(patient_ids, by = "sample_name")

# Save patient_id and visit_id order
annotation_data <- data.frame(visit_id = patient_data$visit_id)
rownames(annotation_data) <- paste0(patient_data$patient_id, "_batch", patient_data$batch, "_", patient_data$visit_id)
rownames(patient_data) <- rownames(annotation_data)
```

#Heatmap for All Data
```{r gen_heatmaps, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

# Used mainly to look and see if there's any batch effects or other things 
# (may not be needed since all visits are included)

if(!dir.exists(paste0(working_dir, "heatmap_by_order"))){
  dir.create(paste0(working_dir, "heatmap_by_order"), recursive = TRUE)
}

# Arrange data by id
annotation_data <- patient_data %>% arrange(order) %>% 
                          select(batch, visit_id) %>% 
                          mutate(batch = as.factor(batch),
                                 visit_id = as.factor(visit_id))
rownames(annotation_data) <- patient_data %>% arrange(order) %>% pull(id)

px_list <- patient_data %>% arrange(order) %>% pull(patient_id)
px_gaps <- which(px_list[-1] != px_list[-length(px_list)])

plot_data <- patient_data %>% arrange(order) %>% 
                          select(!c(sample_name, cohort, cartridge, patient_id, donor_id, visit_id, timepoint, batch, set, order, id, hours)) %>% 
                          mutate(across(everything(), as.numeric)) %>% 
                          as.matrix()
rownames(plot_data) <- rownames(annotation_data)

temp_plot <- pheatmap(plot_data,
                      annotation_row = annotation_data,
                      cluster_rows = FALSE, # Preserve patient order
                      show_rownames = TRUE,
                      show_colnames = TRUE,
                      gaps_row = px_gaps,
                      scale = "row",
                      color = colorRampPalette(c("darkblue", "white", "darkred"))(100),
                      main = paste0("Log2FC for all genes Signature"))

temp_plot

png(filename = paste0(working_dir, "heatmap_by_order/heatmap_log2fc_allgenes_by_id.png"), 
        width = 1600, height = 1000, res = 150)
    grid::grid.newpage()
    grid::grid.draw(temp_plot$gtable)
dev.off()

# Arrange data by batch
annotation_data <- patient_data %>% arrange(batch) %>% filter(patient_id != "control") %>%
                          select(batch, visit_id) %>% 
                          mutate(batch = as.factor(batch),
                                 visit_id = as.factor(visit_id))
rownames(annotation_data) <- patient_data %>% arrange(batch) %>% filter(patient_id != "control") %>% pull(id)

px_list <- patient_data %>% arrange(batch) %>% filter(patient_id != "control") %>% pull(batch)
px_gaps <- which(px_list[-1] != px_list[-length(px_list)])

plot_data <- patient_data %>% arrange(batch) %>% filter(patient_id != "control") %>% 
                          select(!c(sample_name, cohort, cartridge, patient_id, donor_id, visit_id, timepoint, batch, set, order, id)) %>% 
                          mutate(across(everything(), as.numeric)) %>% 
                          as.matrix()
rownames(plot_data) <- rownames(annotation_data)

temp_plot <- pheatmap(plot_data,
                      annotation_row = annotation_data,
                      cluster_rows = FALSE, # Preserve patient order
                      show_rownames = TRUE,
                      show_colnames = TRUE,
                      gaps_row = px_gaps,
                      scale = "row",
                      color = colorRampPalette(c("darkblue", "white", "darkred"))(100),
                      main = paste0("Log2FC for all genes Signature"))

temp_plot

png(filename = paste0(working_dir, "heatmap_by_order/heatmap_log2fc_allgenes_by_batch.png"), 
        width = 1600, height = 1000, res = 150)
    grid::grid.newpage()
    grid::grid.draw(temp_plot$gtable)
dev.off()
```

#Heatmaps, Nanostring
```{r gen_heatmaps, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

# Start with NanoString annotations
for(probeset in ns_probe_list){
  probeset_labels <- ns_probe_annotations %>% 
    filter(grepl(probeset, Related.Probes)) %>%
    pull(Codeset.Name)
  
  # Arrange data by visit_id 
  plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control") %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                    "visit_id", "id", "cohort", "patient_id", "order")))
  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "nanostring",
                    output_path = working_dir)
}
```

#Heatmaps, Custom
```{r gen_heatmaps_custom, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

# Use custom annotations
for(sig_num in 1:length(signature_list)){
  
  probeset_labels <- signature_list[[sig_num]]
  probeset <- names(signature_list)[sig_num]
  
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  # Arrange data by visit_id 
   plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control") %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                    "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "custom",
                    output_path = working_dir)
}
```

#Heatmaps for Cohorts 2-3
```{r gen_heatmaps_custom_2-3, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

# Use custom annotations
for(sig_num in 1:length(signature_list)){
  
  probeset_labels <- signature_list[[sig_num]]
  probeset <- names(signature_list)[sig_num]
  
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  # Plot all data
  # Arrange data by visit_id 
   plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1") %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                    "visit_id", "id", "cohort", "patient_id", "order")))

    # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "custom",
                    output_path = paste0(working_dir, "cohort2-3/"))
  
  if(!dir.exists(paste0(working_dir, "cohort2-3/csv/"))){
      dir.create(paste0(working_dir, "cohort2-3/csv/"), recursive = TRUE)
  }
  
  write.csv(x = temp_data,
            file = paste0(working_dir, "cohort2-3/csv/log2fc_baseline_custom_", probeset, "_all.csv"))
  
  # Plot first 24 hours
  plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1" & visit_id %in% c("baseline", "D001_H01", "D001_H04", "D001_H12", "D001_H24")) %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                  "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "custom_pre24",
                    output_path = paste0(working_dir, "cohort2-3/"))

  # Plot after 24 hours
  plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1" & !visit_id %in% c("D001_H01", "D001_H04", "D001_H12", "D001_H24")) %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                  "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "custom_post24",
                    output_path = paste0(working_dir, "cohort2-3/"))
}

# Start with NanoString annotations
for(probeset in ns_probe_list){
  probeset_labels <- ns_probe_annotations %>% 
    filter(grepl(probeset, Related.Probes)) %>%
    pull(Codeset.Name)
  
  # Plot all data
  # Arrange data by visit_id 
   plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1") %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                    "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "nanostring",
                    output_path = paste0(working_dir, "cohort2-3/"))
  
  write.csv(x = temp_data,
          file = paste0(working_dir, "cohort2-3/csv/log2fc_baseline_nanostring_", probeset, "_all.csv"))
  
  # Plot first 24 hours
  plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1" & visit_id %in% c("baseline", "D001_H01", "D001_H04", "D001_H12", "D001_H24")) %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                  "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "nanostring_pre24",
                    output_path = paste0(working_dir, "cohort2-3/"))
  
  # Plot after 24 hours
  plot_data <- patient_data %>% 
    arrange(visit_id) %>% 
    filter(patient_id != "control" & cohort != "1" & !visit_id %in% c("D001_H01", "D001_H04", "D001_H12", "D001_H24")) %>%
    as.data.frame() %>%
    select(any_of(c(probeset_labels,
                  "visit_id", "id", "cohort", "patient_id", "order")))

  
  # Add an identifier 'patient' a combination of patient_id and cohort
  plot_data$patient <- paste0(plot_data$patient_id, "_", plot_data$cohort)
  
  temp_data <- make_gene_heatmap(probeset_labels = probeset_labels,
                    probeset = probeset,
                    baseline_visit = "baseline",
                    plot_data = plot_data,
                    file_id = "nanostring_post24",
                    output_path = paste0(working_dir, "cohort2-3/"))
}
```

#Visit Dotplots, All
```{r gen_dotplots, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 5}

#NanoString signatures
for(probeset in ns_probe_list){
  probeset_labels <- ns_probe_annotations %>% 
    filter(grepl(probeset, Related.Probes)) %>%
    pull(Codeset.Name)
  
  plot_data <- patient_data %>% 
                  arrange(order) %>% 
                  filter(patient_id != "control") %>%
                  select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint")))
  
  plot_data$id <- sub("_.*", "", plot_data$id)
  
  plot_ids <- unique(plot_data$id)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)
  
  plot_cohorts <- sort(unique(plot_data$cohort))
  color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
  color_map <- setNames(color_values, plot_cohorts)
    
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  make_gene_dotplot(plot_data = plot_data,
                    probeset = probeset,
                    probeset_labels = probeset_labels,
                    file_id = "nanostring")
  
  make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
    file_id = "nanostring_cohort1")
    
  make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "2"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
    file_id = "nanostring_cohort2")
  
    make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "3"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
    file_id = "nanostring_cohort3")
    
    make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort != "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort != "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort != "1") %>% pull(cohort)))],
    file_id = "nanostring_cohort2-3")
}

# Custom signatures
for(sig_num in 1:length(signature_list)){

  probeset_labels <- signature_list[[sig_num]]
  probeset <- names(signature_list)[sig_num]
  
  plot_data <- patient_data %>% 
                  arrange(order) %>% 
                  filter(patient_id != "control") %>%
                  select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint")))
  
  plot_data$id <- sub("_.*", "", plot_data$id)
  
  plot_ids <- unique(plot_data$id)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)
  
  plot_cohorts <- sort(unique(plot_data$cohort))
  color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
  color_map <- setNames(color_values, plot_cohorts)
    
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  make_gene_dotplot(plot_data = plot_data,
                    probeset = probeset,
                    probeset_labels = probeset_labels,
                    file_id = "custom")
  
  make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
    file_id = "custom_cohort1")
    
  make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "2"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
    file_id = "custom_cohort2")
  
    make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "3"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
    file_id = "custom_cohort3")
    
    make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort != "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort != "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort != "1") %>% pull(cohort)))],
    file_id = "custom_cohort2-3")
}
```

#Patient C2P1 Dotplots, All
```{r gen_dotplots_c2p1, include = FALSE, echo = FALSE, fig.width = 10, fig.height = 5}

ns_probesets <- list(
  "Complement" = ns_probe_annotations %>% filter(grepl("Complement", Related.Probes)) %>% pull(Codeset.Name), 
  "Adhesion" = ns_probe_annotations %>% filter(grepl("Adhesion", Related.Probes)) %>% pull(Codeset.Name),
  "T-Cell Functions" = ns_probe_annotations %>% filter(grepl("T-Cell Functions", Related.Probes)) %>% pull(Codeset.Name),
  "Interleukins" = ns_probe_annotations %>% filter(grepl("Interleukins", Related.Probes)) %>% pull(Codeset.Name),
  "Chemokines" = ns_probe_annotations %>% filter(grepl("Chemokines", Related.Probes)) %>% pull(Codeset.Name),
  "Cytokines" = ns_probe_annotations %>% filter(grepl("Cytokines", Related.Probes)) %>% pull(Codeset.Name),
  "Regulation" = ns_probe_annotations %>% filter(grepl("Regulation", Related.Probes)) %>% pull(Codeset.Name),
  "NK Cell Functions" = ns_probe_annotations %>% filter(grepl("NK Cell Functions", Related.Probes)) %>% pull(Codeset.Name),
  "TLR" = ns_probe_annotations %>% filter(grepl("TLR", Related.Probes)) %>% pull(Codeset.Name),
  "CT Antigen" = ns_probe_annotations %>% filter(grepl("CT Antigen", Related.Probes)) %>% pull(Codeset.Name),
  "Cell Cycle" = ns_probe_annotations %>% filter(grepl("Cell Cycle", Related.Probes)) %>% pull(Codeset.Name),
  "Antigen Processing" = ns_probe_annotations %>% filter(grepl("Antigen Processing", Related.Probes)) %>% pull(Codeset.Name),
  "TNF Superfamily" = ns_probe_annotations %>% filter(grepl("TNF Superfamily", Related.Probes)) %>% pull(Codeset.Name),
  "Pathogen Defense" = ns_probe_annotations %>% filter(grepl("Pathogen Defense", Related.Probes)) %>% pull(Codeset.Name),
  "Cell Functions" = ns_probe_annotations %>% filter(grepl("Cell Functions", Related.Probes)) %>% pull(Codeset.Name),
  "Transporter Functions" = ns_probe_annotations %>% filter(grepl("Transporter Functions", Related.Probes)) %>% pull(Codeset.Name),
  "B-Cell Functions" = ns_probe_annotations %>% filter(grepl("B-Cell Functions", Related.Probes)) %>% pull(Codeset.Name),
  "Senescence" = ns_probe_annotations %>% filter(grepl("Senescence", Related.Probes)) %>% pull(Codeset.Name),
  "Leukocyte Functions" = ns_probe_annotations %>% filter(grepl("Leukocyte Functions", Related.Probes)) %>% pull(Codeset.Name),
  "Microglial Functions" = ns_probe_annotations %>% filter(grepl("Microglial Functions", Related.Probes)) %>% pull(Codeset.Name),
  "Cytotoxicity" = ns_probe_annotations %>% filter(grepl("Cytotoxicity", Related.Probes)) %>% pull(Codeset.Name)
)

patient <- "100-001"

all_probesets <- c(ns_probesets, signature_list)

temp_data <- patient_data %>% arrange(visit_id) %>% select(!c("sample_name", "cohort", "patient_id",
                                                              "visit_id", "order", "id"))

meta_data <- patient_data %>% arrange(visit_id) %>% select(c("sample_name", "cohort", "patient_id",
                                                              "visit_id", "order", "id"))
meta_data$sample_name <- rownames(meta_data)

sample_id <- which(meta_data$patient_id %in% patient)

temp_data <- temp_data[sample_id, , drop = FALSE]

meta_sub <- meta_data %>%
    filter(patient_id == patient) %>%
    arrange(factor(visit_id, levels = unique(visit_id)))

plot_data <- meta_data %>% filter(patient_id == patient) %>%
    arrange(factor(visit_id, levels = unique(visit_id)))

plot_data <- expand.grid(sample = meta_sub$sample_name,
                           signature = names(all_probesets),
                           stringsAsFactors = FALSE) %>%
    left_join(meta_sub %>% select(sample_name, visit_id),
              by = c("sample" = "sample_name")) %>%
    rowwise() 

plot_data$geo_mean <- ""

# Inefficient but rushed
for(row in 1:nrow(plot_data)){
  probes <- all_probesets[[plot_data$signature[row]]]
  probes <- probes[probes %in% colnames(temp_data)]
  geo_mean <- temp_data[plot_data$sample[row], ] %>% select(any_of(probes)) %>% calc_geom_mean()
  plot_data$geo_mean[row] <- geo_mean
}

plot_data$geo_mean <- as.numeric(plot_data$geo_mean)

# Plot

n_cats <- length(unique(plot_data$signature))
colours_34 <- randomcoloR::distinctColorPalette(n_cats)
shapes_34  <- rep(0:25, length.out = n_cats)

sigs <- plot_data %>% filter(signature %in% names(ns_probesets)) %>% pull(signature) %>% unique() %>% sort()
colours_34[which(sigs %in% "Adhesion")] <- "black"
shapes_34[which(sigs %in% "Adhesion")] <- 16

plotted_data <- ggplot(plot_data %>% filter(signature %in% names(ns_probesets)), aes(x = visit_id, y = geo_mean, group = signature)) +
    geom_line(aes(color = factor(signature))) +
    geom_point(aes(color = factor(signature), shape = signature), size = 3) +
    theme_bw() +
    labs(title = paste0("C2P1 signatures over visit"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    ylim(min(plot_data$geo_mean) - 1, max(plot_data$geo_mean) + 1) +
    theme(legend.position = "right") +
    scale_color_manual(
      values = colours_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    ) +
    scale_shape_manual(
      values = shapes_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    )

  plot(plotted_data)
  
  png(filename = paste0(working_dir, "dotplot_by_visit/dotplot_C2P1_nanostring_by_visit.png"),
      width = 1000, height = 1000, res = 150)
  plot(plotted_data)
  dev.off()
  
  n_cats <- length(unique(plot_data$signature))
  colours_34 <- randomcoloR::distinctColorPalette(n_cats)
  shapes_34  <- rep(0:25, length.out = n_cats)
  
  sigs <- plot_data %>% filter(signature %in% names(signature_list)) %>% pull(signature) %>% unique() %>% sort()
  colours_34[which(sigs %in% "Type 1 IFN cells")] <- "black"
  shapes_34[which(sigs %in% "Type 1 IFN cells")] <- 16

  plotted_data <- ggplot(plot_data %>% filter(signature %in% names(signature_list)), aes(x = visit_id, y = geo_mean, group = signature)) +
    geom_line(aes(color = factor(signature))) +
    geom_point(aes(color = factor(signature), shape = signature), size = 3) +
    theme_bw() +
    labs(title = paste0("C2P1 signatures over visit"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    ylim(min(plot_data$geo_mean) - 1, max(plot_data$geo_mean) + 1) +
    theme(legend.position = "right") +
    scale_color_manual(
      values = colours_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    ) +
    scale_shape_manual(
      values = shapes_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    )

  plot(plotted_data)
  
  png(filename = paste0(working_dir, "dotplot_by_visit/dotplot_C2P1_custom_by_visit.png"),
      width = 1000, height = 1000, res = 150)
  plot(plotted_data)
  dev.off()
  
  n_cats <- length(unique(plot_data$signature))
  colours_34 <- randomcoloR::distinctColorPalette(n_cats)
  shapes_34  <- rep(0:25, length.out = n_cats)
  
  sigs <- plot_data %>% filter(signature %in% names(signature_list_2)) %>% pull(signature) %>% unique() %>% sort()
  colours_34[which(sigs %in% "Type 1 IFN cells")] <- "black"
  shapes_34[which(sigs %in% "Type 1 IFN cells")] <- 16
  
  plotted_data <- ggplot(plot_data %>% filter(signature %in% names(signature_list_2)), aes(x = visit_id, y = geo_mean, group = signature)) +
    geom_line(aes(color = factor(signature))) +
    geom_point(aes(color = factor(signature), shape = signature), size = 3) +
    theme_bw() +
    labs(title = paste0("C2P1 signatures over visit"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    ylim(min(plot_data$geo_mean) - 1, max(plot_data$geo_mean) + 1) +
    theme(legend.position = "right") +
    scale_color_manual(
      values = colours_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    ) +
    scale_shape_manual(
      values = shapes_34,
      guide  = guide_legend(ncol = 2, override.aes = list(size = 4))
    )

  plot(plotted_data)
  
  png(filename = paste0(working_dir, "dotplot_by_visit/dotplot_C2P1_custom_reduced_by_visit.png"),
      width = 1000, height = 1000, res = 150)
  plot(plotted_data)
  dev.off()


```

#Cohort Custom Dotplots, All
```{r gen_dotplots_custom, include = FALSE, echo = FALSE, fig.width = 10, fig.height = 5}

probeset <- "Type 1 IFN cells"
probeset_labels <- signature_list[[probeset]]
  
plot_data <- patient_data %>% 
                  arrange(order) %>% 
                  filter(patient_id != "control") %>%
                  select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint")))
  
plot_data$id <- sub("_.*", "", plot_data$id)

plot_ids <- unique(plot_data$id)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(plot_data$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)
  
# Remove any probes that aren't in the dataset
probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "1"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
  file_id = "cohort1")
  
make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "2"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
  file_id = "cohort2")

make_gene_custom_dotplot(plot_data = plot_data %>% filter(cohort == "3"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
  file_id = "cohort3")

make_gene_custom_dotplot(plot_data = plot_data,
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data) %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data) %>% pull(cohort)))],
  file_id = "cohort_all")
```

#Cohort Custom Dotplots, Normalized to Baseline
```{r gen_dotplots_custom_baseline_norm, include = FALSE, echo = FALSE, fig.width = 10, fig.height = 5}

make_gene_custom_dotplot_baseline_norm <- function(plot_data, probeset, probeset_labels, file_id, shape_map, color_map){
  
  # Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
  plot_data$geo_mean <- apply(plot_data %>%
                          mutate(across(all_of(probeset_labels), as.numeric)) %>%
                          select(any_of(probeset_labels)) %>%
                          as.matrix(),
                          1, calc_geom_mean)
  
  plot_data <- plot_data %>% 
    select(visit_id, patient_id, id, cohort, timepoint, geo_mean) %>% 
    group_by(patient_id) %>% 
    mutate(baseline_geo = geo_mean[visit_id == "baseline"]) %>% 
    mutate(ratio = geo_mean / baseline_geo)
  
  plot_order <- plot_data %>% arrange(timepoint) %>% pull(visit_id) %>% unique()
  
  plot_data <- plot_data %>%
    mutate(visit_id = factor(visit_id, levels = plot_order)) %>% 
    mutate(cohort = as.numeric(cohort)) %>%
    mutate(ratio = as.numeric(ratio))
  
  # Plot
  plotted_data <- ggplot(plot_data, aes(x = visit_id, y = ratio)) +
    geom_line(aes(group = id, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = id, color = factor(cohort)), size = 4, stroke = 1.2, fill = "white") +
    theme_bw() +
    scale_shape_manual(name = "id", values = shape_map) + 
    scale_color_manual(name = "cohort", values = color_map) +
    labs(title = paste0(probeset, " signature over visit_id by patient_id"),
         x = "Visit",
         y = "Log2 Geo Mean Ratio to Baseline",
         color = "cohort") +
    ylim(min(plot_data$ratio) - .5, max(plot_data$ratio) + .5) +
    theme(legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1))

  plot(plotted_data)
  
  if(!dir.exists(paste0(working_dir, "dotplot_by_visit/baseline_ratio"))){
    dir.create(paste0(working_dir, "dotplot_by_visit/baseline_ratio"), recursive = TRUE)
  }
  
  png(filename = paste0(working_dir, "dotplot_by_visit/baseline_ratio/dotplot_", file_id, "_", probeset, ".png"))
  plot(plotted_data)
  dev.off()
}

probeset <- "Type 1 IFN cells"
probeset_labels <- signature_list[[probeset]]
  
plot_data <- patient_data %>% 
                  arrange(order) %>% 
                  filter(patient_id != "control") %>%
                  select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint")))
  
plot_data$id <- sub("_.*", "", plot_data$id)

plot_ids <- unique(plot_data$id)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(plot_data$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)
  
# Remove any probes that aren't in the dataset
probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
make_gene_custom_dotplot_baseline_norm(plot_data = plot_data %>% filter(cohort == "1"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
  file_id = "cohort1")
  
make_gene_custom_dotplot_baseline_norm(plot_data = plot_data %>% filter(cohort == "2"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
  file_id = "cohort2")

make_gene_custom_dotplot_baseline_norm(plot_data = plot_data %>% filter(cohort == "3"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
  file_id = "cohort3")

make_gene_custom_dotplot_baseline_norm(plot_data = plot_data,
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% pull(cohort)))],
  file_id = "cohort_all")
```

#Cohort Custom Dotplots, Hours
```{r gen_dotplots_custom_hours, include = FALSE, echo = FALSE, fig.width = 20, fig.height = 5}

make_gene_custom_dotplot_hours <- function(plot_data, probeset, probeset_labels, file_id, shape_map, color_map, xmin, xmax, ymin, ymax){
  
  plot_order <- plot_data %>% arrange(timepoint) %>% pull(hours) %>% unique()
  
  plot_data <- plot_data %>%
    mutate(cohort = as.numeric(cohort))
  
  # Plots
  plotted_data <- ggplot(plot_data, aes(x = hours, y = geo_mean)) +
    geom_line(aes(group = id, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = id, color = factor(cohort)), size = 4, stroke = 1.2, fill = "white") +
    theme_bw() +
    scale_shape_manual(name = "id", values = shape_map) + 
    scale_color_manual(name = "cohort", values = color_map) +
    scale_x_continuous(
      name = "Hours Post-treatment",
      limits = c(xmin, xmax),
      breaks = seq(xmin, xmax, by = 24) # 24 hour breakpoints
    ) +
    coord_cartesian(ylim = c(ymin, ymax)) +
    labs(title = paste0(probeset, " signature over hours by patient_id"),
         x = "Visit",
         y = "Log2 Geo Mean",
         color = "cohort") +
    theme(legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1))

  plot(plotted_data)
  
  if(!dir.exists(paste0(working_dir, "dotplot_by_visit/hours"))){
    dir.create(paste0(working_dir, "dotplot_by_visit/hours"), recursive = TRUE)
  }
  
  png(filename = paste0(working_dir, "dotplot_by_visit/hours/dotplot_hours_", file_id, "_", probeset, ".png"), 
      width = 1600, height = 400, units = "px")
  plot(plotted_data)
  dev.off()
}

# Some custom analysis for specific cohorts
probeset <- "Type 1 IFN cells"
probeset_labels <- signature_list[[probeset]]
  
plot_data <- patient_data %>% 
                  arrange(order) %>% 
                  filter(patient_id != "control") %>%
                  select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint", "hours")))
  
plot_data$id <- sub("_.*", "", plot_data$id)

plot_data <- plot_data %>% mutate(hours = as.numeric(hours))

# Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
plot_data$geo_mean <- apply(plot_data %>%
  mutate(across(all_of(probeset_labels), as.numeric)) %>%
  select(any_of(probeset_labels)) %>%
  as.matrix(),
  1, calc_geom_mean)

ymin <- min(plot_data$geo_mean, na.rm=TRUE) - 1
ymax <- max(plot_data$geo_mean, na.rm=TRUE) + 1
xmin <- 0
xmax <- plot_data$hours %>% max()

plot_ids <- unique(plot_data$id)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(plot_data$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)
  
# Remove any probes that aren't in the dataset
probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
# Make a custom plot to look at xmin/xmax where Cohort 3 cutoff is

xmax <- plot_data %>% filter(cohort == "3") %>% pull(hours) %>% max()

make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "1"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
  file_id = "cohort1_cohort3_scale", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
  
make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "2"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
  file_id = "cohort2_cohort3_scale", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)

make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "3"),
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
  file_id = "cohort3_cohort3_scale", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)

make_gene_custom_dotplot_hours(plot_data = plot_data,
  probeset = probeset,
  probeset_labels = probeset_labels,
  shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% pull(id)))],
  color_map = color_map[which(names(color_map) %in% unique(plot_data %>% pull(cohort)))],
  file_id = "cohort_all_cohort3_scale", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)

# Cycle through all the custom signatures
for(sig_num in 1:length(signature_list)){

  probeset_labels <- signature_list[[sig_num]]
  probeset <- names(signature_list)[sig_num]

  plot_data <- patient_data %>% 
                    arrange(order) %>% 
                    filter(patient_id != "control") %>%
                    select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint", "hours")))
    
  plot_data$id <- sub("_.*", "", plot_data$id)
  
  plot_data <- plot_data %>% mutate(hours = as.numeric(hours))
  
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  # Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
  plot_data$geo_mean <- apply(plot_data %>%
    mutate(across(all_of(probeset_labels), as.numeric)) %>%
    select(any_of(probeset_labels)) %>%
    as.matrix(),
    1, calc_geom_mean)
  
  ymin <- min(plot_data$geo_mean, na.rm=TRUE) - 1
  ymax <- max(plot_data$geo_mean, na.rm=TRUE) + 1
  xmin <- 0
  xmax <- plot_data$hours %>% max()
  
  plot_ids <- unique(plot_data$id)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)
    
  plot_cohorts <- sort(unique(plot_data$cohort))
  color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
  color_map <- setNames(color_values, plot_cohorts)
    
  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
    file_id = "cohort1", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
    
  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "2"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
    file_id = "cohort2", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
  
  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "3"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
    file_id = "cohort3", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
  
  make_gene_custom_dotplot_hours(plot_data = plot_data,
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% pull(cohort)))],
    file_id = "cohort_all", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
}

# Cycle through all the nanostring signatures
for(probeset in ns_probe_list){
  
  probeset_labels <- ns_probe_annotations %>% 
    filter(grepl(probeset, Related.Probes)) %>%
    pull(Codeset.Name)

  plot_data <- patient_data %>% 
                    arrange(order) %>% 
                    filter(patient_id != "control") %>%
                    select(any_of(c(probeset_labels, "visit_id", "id", "patient_id", "cohort", "timepoint", "hours")))
    
  plot_data$id <- sub("_.*", "", plot_data$id)
  
  plot_data <- plot_data %>% mutate(hours = as.numeric(hours))
  
  # Remove any probes that aren't in the dataset
  probeset_labels <- probeset_labels[probeset_labels %in% colnames(patient_data)]
  
  # Gene data is in columns 1 through n (the last columns will be visit id and id) in plot_data:
  plot_data$geo_mean <- apply(plot_data %>%
    mutate(across(all_of(probeset_labels), as.numeric)) %>%
    select(any_of(probeset_labels)) %>%
    as.matrix(),
    1, calc_geom_mean)
  
  ymin <- min(plot_data$geo_mean, na.rm=TRUE) - 1
  ymax <- max(plot_data$geo_mean, na.rm=TRUE) + 1
  xmin <- 0
  xmax <- plot_data$hours %>% max()
  
  plot_ids <- unique(plot_data$id)
  shape_codes <- c(1:25)[seq_along(plot_ids)]
  shape_map <- setNames(shape_codes, plot_ids)
    
  plot_cohorts <- sort(unique(plot_data$cohort))
  color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
  color_map <- setNames(color_values, plot_cohorts)

  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "1"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "1") %>% pull(cohort)))],
    file_id = "cohort1", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
    
  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "2"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "2") %>% pull(cohort)))],
    file_id = "cohort2", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
  
  make_gene_custom_dotplot_hours(plot_data = plot_data %>% filter(cohort == "3"),
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% filter(cohort == "3") %>% pull(cohort)))],
    file_id = "cohort3", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
  
  make_gene_custom_dotplot_hours(plot_data = plot_data,
    probeset = probeset,
    probeset_labels = probeset_labels,
    shape_map = shape_map[which(names(shape_map) %in% unique(plot_data %>% pull(id)))],
    color_map = color_map[which(names(color_map) %in% unique(plot_data %>% pull(cohort)))],
    file_id = "cohort_all", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)
}
```

#Cell Profiler Module Results
```{r gen_cellprofiler_plots, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 5}

# Customer request:
# The customer wants to re-plot the relative and raw enrichment scores for the 
# Cell Profiler Module in NanoString.

# Some reformatting of the file to fit in with existing workflow
raw_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_raw.csv", 
                                   skip = 1)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "X"] <- "sample_name"
raw_cellprofiler_input$sample_name <- gsub("-", ".", raw_cellprofiler_input$sample_name)
raw_cellprofiler_input <- raw_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
raw_cellprofiler_input$px <- sub("_.*", "", raw_cellprofiler_input$id)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "visit_id"] <- "tp"
raw_cellprofiler_input$sample_name <- NULL

relative_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_relative.csv",
                                   skip = 1)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "X"] <- "sample_name"
relative_cellprofiler_input$sample_name <- gsub("-", ".", relative_cellprofiler_input$sample_name)
relative_cellprofiler_input <- relative_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
relative_cellprofiler_input$px <- sub("_.*", "", relative_cellprofiler_input$id)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "visit_id"] <- "tp"
relative_cellprofiler_input$sample_name <- NULL

timepoint_levels <- patient_ids %>% filter(patient_id != "control") %>% arrange(timepoint) %>% pull(visit_id) %>% unique()

celltypes <- raw_cellprofiler_input %>% select(-c(id,px,tp,cohort)) %>% colnames()

for(celltype in celltypes){
  plot_celltype_score(cellprofiler_input = raw_cellprofiler_input,
                      celltype = celltype,
                      data_type = "raw_enrichment",
                      timepoint_levels = timepoint_levels)
}


celltypes <- relative_cellprofiler_input %>% select(-c(id,px,tp,cohort)) %>% colnames()

for(celltype in celltypes){
  plot_celltype_score(cellprofiler_input = relative_cellprofiler_input,
                      celltype = celltype,
                      data_type = "relative_enrichment",
                      timepoint_levels = timepoint_levels)
}

```

# Cell Profiler, Individual Cohort Combinations
```{r gen_cellprofiler_plots_custom, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 5}

# Grab only data from cohort 3
raw_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_raw.csv", 
                                   skip = 1)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "X"] <- "sample_name"
raw_cellprofiler_input$sample_name <- gsub("-", ".", raw_cellprofiler_input$sample_name)
raw_cellprofiler_input <- raw_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
raw_cellprofiler_input$px <- sub("_.*", "", raw_cellprofiler_input$id)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "visit_id"] <- "tp"
raw_cellprofiler_input$sample_name <- NULL

relative_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_relative.csv",
                                   skip = 1)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "X"] <- "sample_name"
relative_cellprofiler_input$sample_name <- gsub("-", ".", relative_cellprofiler_input$sample_name)
relative_cellprofiler_input <- relative_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
relative_cellprofiler_input$px <- sub("_.*", "", relative_cellprofiler_input$id)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "visit_id"] <- "tp"
relative_cellprofiler_input$sample_name <- NULL

# New function for custom plotting
# Quick and dirty way to plot individual cell profiler scores
plot_celltype_score_custom  <- function (cellprofiler_input, celltype, data_type, timepoint_levels, output_path, shape_map, color_map){
  
  celltype_score <- cellprofiler_input %>%
    pivot_longer(
      cols = -c(id, cohort, px, tp),
      names_to  = "cell_type",
      values_to = "abundance"
    ) %>%
    mutate(tp = factor(tp, levels = timepoint_levels))
  
  celltype_score$abundance <- as.numeric(celltype_score$abundance)
  
  temp_plot <- ggplot(celltype_score %>% filter(cell_type == celltype), aes(x = tp, y = abundance)) +
    geom_line(aes(group = px, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = px, color = factor(cohort)),
               size = 4, stroke = 1.2, fill = "white") +
    scale_shape_manual(name   = "Sample ID",
                       values = shape_map) +
    scale_color_manual(name = "Cohort", values = color_map) +
    theme_minimal() +
    theme(
      axis.text.x  = element_text(angle = 30, hjust = 1),
      legend.title = element_text(face = "bold")
    ) +
    ylim(min(celltype_score %>% filter(cell_type == celltype) %>% select(abundance)) - 1, 
         max(celltype_score %>% filter(cell_type == celltype) %>% select(abundance)) + 1) +
    labs(
      title = paste0(celltype, " Abundance Over Treatment (", data_type, ")"),
      x     = "Timepoint",
      y     = "Estimated Abundance"
    )
  
  plot(temp_plot)
  
  if(!dir.exists(output_path)){
      dir.create(output_path, recursive = TRUE)
  }
  
  png(filename = file.path(output_path, paste0(celltype, "_", data_type, "_cellprofiler_score.png")), 
        width = 1000, height = 1000, res = 150)
  
  plot(temp_plot)
  
  dev.off()
}

celltypes <- raw_cellprofiler_input %>% select(-c(id,px,tp,cohort)) %>% colnames()
timepoint_levels <- patient_ids %>% filter(patient_id != "control") %>% arrange(timepoint) %>% pull(visit_id) %>% unique()

plot_ids <- unique(raw_cellprofiler_input$px)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(raw_cellprofiler_input$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)

for(celltype in celltypes){
  plot_celltype_score_custom(cellprofiler_input = raw_cellprofiler_input %>% filter(cohort == "3"),
                      celltype = celltype,
                      data_type = "raw_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "3") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "3") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort3"))
  
  plot_celltype_score_custom(cellprofiler_input = raw_cellprofiler_input %>% filter(cohort == "2"),
                      celltype = celltype,
                      data_type = "raw_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "2") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "2") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort2"))
  
  plot_celltype_score_custom(cellprofiler_input = raw_cellprofiler_input %>% filter(cohort != "1"),
                      celltype = celltype,
                      data_type = "raw_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(raw_cellprofiler_input %>% filter(cohort != "1") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(raw_cellprofiler_input %>% filter(cohort != "1") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort2-3"))
}


celltypes <- relative_cellprofiler_input %>% select(-c(id,px,tp,cohort)) %>% colnames()

plot_ids <- unique(relative_cellprofiler_input$px)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(relative_cellprofiler_input$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)

for(celltype in celltypes){
  plot_celltype_score_custom(cellprofiler_input = relative_cellprofiler_input %>% filter(cohort == "3"),
                      celltype = celltype,
                      data_type = "relative_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(relative_cellprofiler_input %>% filter(cohort == "3") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(relative_cellprofiler_input %>% filter(cohort == "3") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort3"))
  
  plot_celltype_score_custom(cellprofiler_input = relative_cellprofiler_input %>% filter(cohort == "2"),
                      celltype = celltype,
                      data_type = "relative_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(relative_cellprofiler_input %>% filter(cohort == "2") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(relative_cellprofiler_input %>% filter(cohort == "2") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort2"))
  
  plot_celltype_score_custom(cellprofiler_input = relative_cellprofiler_input %>% filter(cohort != "1"),
                      celltype = celltype,
                      data_type = "relative_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(relative_cellprofiler_input %>% filter(cohort != "1") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(relative_cellprofiler_input %>% filter(cohort != "1") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort2-3"))
}

```

#Cell Profiler, Custom Neutrophil Removal
```{r gen_cellprofiler_plots_custom, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 5}

# Grab only data from cohort 3
raw_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_raw.csv", 
                                   skip = 1)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "X"] <- "sample_name"
raw_cellprofiler_input$sample_name <- gsub("-", ".", raw_cellprofiler_input$sample_name)
raw_cellprofiler_input <- raw_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
raw_cellprofiler_input$px <- sub("_.*", "", raw_cellprofiler_input$id)
colnames(raw_cellprofiler_input)[colnames(raw_cellprofiler_input) == "visit_id"] <- "tp"
raw_cellprofiler_input$sample_name <- NULL

relative_cellprofiler_input <- read.csv("/home/byuen/projects/actymthera/projects/ns_cohort_1-3/raw_data/celltype_score_relative.csv",
                                   skip = 1)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "X"] <- "sample_name"
relative_cellprofiler_input$sample_name <- gsub("-", ".", relative_cellprofiler_input$sample_name)
relative_cellprofiler_input <- relative_cellprofiler_input %>% 
                            dplyr::inner_join(patient_ids, by = "sample_name") %>%
                            select(-c(order, set, batch, timepoint, cartridge, patient_id, donor_id, hours))
relative_cellprofiler_input$px <- sub("_.*", "", relative_cellprofiler_input$id)
colnames(relative_cellprofiler_input)[colnames(relative_cellprofiler_input) == "visit_id"] <- "tp"
relative_cellprofiler_input$sample_name <- NULL

# New function for custom plotting
# Quick and dirty way to plot individual cell profiler scores
plot_celltype_score_custom  <- function (cellprofiler_input, celltype, data_type, timepoint_levels, output_path, shape_map, color_map){
  
  # Generate TIL score that includes neutrophil
  cellprofiler_input$TIL.with.Neutrophil <- (cellprofiler_input$B.cells + cellprofiler_input$CD45 + 
                                               cellprofiler_input$Cytotoxic.cells + cellprofiler_input$Macrophages +
                                               cellprofiler_input$T.cells + cellprofiler_input$Neutrophils)/6
  
  adjusted_input <- cellprofiler_input %>% group_by(px, tp) %>% mutate(adj.score = !!sym(celltype) - TIL.with.Neutrophil)
    
  celltype_score <- adjusted_input %>%
    select(cohort, tp, id, px, adj.score) %>%
    pivot_longer(
      cols = -c(id, cohort, px, tp),
      names_to  = "cell_type",
      values_to = "abundance"
    ) %>%
    mutate(tp = factor(tp, levels = timepoint_levels))
  
  celltype_score$abundance <- as.numeric(celltype_score$abundance)
  
  temp_plot <- ggplot(celltype_score, aes(x = tp, y = abundance)) +
    geom_line(aes(group = px, color = factor(cohort)), size = 1) +
    geom_point(aes(shape = px, color = factor(cohort)),
               size = 4, stroke = 1.2, fill = "white") +
    scale_shape_manual(name   = "Sample ID",
                       values = shape_map) +
    scale_color_manual(name = "Cohort", values = color_map) +
    theme_minimal() +
    theme(
      axis.text.x  = element_text(angle = 30, hjust = 1),
      legend.title = element_text(face = "bold")
    ) +
    ylim(min(celltype_score$abundance) - 1, 
         max(celltype_score$abundance) + 1) +
    labs(
      title = paste0(celltype, " Abundance Over Treatment (", data_type, ")"),
      x     = "Timepoint",
      y     = "Estimated Abundance"
    )
  
  plot(temp_plot)
  
  if(!dir.exists(output_path)){
      dir.create(output_path, recursive = TRUE)
  }
  
  png(filename = file.path(output_path, paste0(celltype, "_", data_type, "_til_neutrophil_cellprofiler_score.png")), 
        width = 1000, height = 1000, res = 150)
  
  plot(temp_plot)
  
  dev.off()
}

celltypes <- raw_cellprofiler_input %>% select(-c(id,px,tp,cohort)) %>% colnames()
celltypes <- celltypes[!celltypes %in% "Neutrophils"] # Remove Neutrophils

timepoint_levels <- patient_ids %>% filter(patient_id != "control") %>% arrange(timepoint) %>% pull(visit_id) %>% unique()

plot_ids <- unique(raw_cellprofiler_input$px)
shape_codes <- c(1:25)[seq_along(plot_ids)]
shape_map <- setNames(shape_codes, plot_ids)
  
plot_cohorts <- sort(unique(raw_cellprofiler_input$cohort))
color_values <- RColorBrewer::brewer.pal(length(plot_cohorts), "Set1")
color_map <- setNames(color_values, plot_cohorts)

for(celltype in celltypes){
  plot_celltype_score_custom(cellprofiler_input = raw_cellprofiler_input %>% filter(cohort == "3"),
                      celltype = celltype,
                      data_type = "custom_enrichment",
                      timepoint_levels = timepoint_levels,
                      shape_map = shape_map[which(names(shape_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "3") %>% pull(px)))],
                      color_map = color_map[which(names(color_map) %in% unique(raw_cellprofiler_input %>% filter(cohort == "3") %>% pull(cohort)))],
                      output_path = paste0(working_dir, "cell_profiler/cell_profiler_cohort3"))
}

```

#Differential Gene Expression and Enrichment, Timepoint vs. Cohort
```{r gen_gene_enrich_timepoint_cohort, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

layer_1 <- "visit_id"
layer_2 <- "cohort"

meta_data <- patient_data %>% filter(patient_id != "control") %>%
              arrange(!!sym(layer_1)) %>% 
              select(c(!!sym(layer_1), !!sym(layer_2)))
meta_data$sample_name <- rownames(meta_data)

# ID the number of samples per cohort visit_id
sample_pool <- meta_data %>% 
  group_by(!!sym(layer_1), !!sym(layer_2)) %>% 
  tally(name = "n_samples")

# ID the visit_id with >=2 samples per visit_id and cohort
samples_to_select <- sample_pool %>%
  filter(n_samples >= 2) %>%
  group_by(!!sym(layer_1)) %>%
  tally(name = "n_groups") %>%
  filter(n_groups >= 2) %>%
  pull(!!sym(layer_1))

# ID the cohort and visit_ids to pull sample_name from
sample_pool <- sample_pool %>%
  filter(!!sym(layer_1) %in% samples_to_select, n_samples >= 2) %>%
  select(!!sym(layer_1), !!sym(layer_2))

# Select sample_name to assess further
samples_to_select <- meta_data %>% 
  semi_join(sample_pool, by = c(layer_1, layer_2)) %>%
  pull(sample_name)

# Partition meta_data to contain viable samples
meta_data <- meta_data %>%
  filter(sample_name %in% samples_to_select)

# Partition expression matrix for samples in question
plot_data <- patient_data %>% 
  filter(patient_id != "control" & rownames(patient_data) %in% samples_to_select) %>%
              arrange(visit_id) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

de_genes <- run_group_by_split_DE(
     expr_mat = plot_data,
     meta = meta_data,
     split_var = layer_1,
     group_var = layer_2
 )

```

#Differential Gene Expression and Enrichment, Cohort vs. Visit
```{r gen_gene_enrich_cohort_timepoint, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

#' Need to run run_group_by_split_DE for timepoints relative to baseline between cohorts
#' @layer_1 should be @cohort so that it can iterate cohort to cohort between timepoints
#' @layer_2 should be @visit_id which should compare all timepoints to each other (including timepoints vs. non-baseline)
#' Will need to create a new @cohort_all layer to do the final comparison of all treatments relative to each timepoint

layer_1 <- "cohort"
layer_2 <- "visit_id"

meta_data <- patient_data %>% filter(patient_id != "control") %>%
              arrange(!!sym(layer_1)) %>% 
              select(c(!!sym(layer_1), !!sym(layer_2)))
meta_data$sample_name <- rownames(meta_data)

# ID the number of samples per cohort visit_id
sample_pool <- meta_data %>% 
  group_by(!!sym(layer_1), !!sym(layer_2)) %>% 
  tally(name = "n_samples")

# ID the visit_id with >=2 samples per visit_id and cohort
samples_to_select <- sample_pool %>%
  filter(n_samples >= 2) %>%
  group_by(!!sym(layer_1)) %>%
  tally(name = "n_groups") %>%
  filter(n_groups >= 2) %>%
  pull(!!sym(layer_1))

# ID the cohort and visit_ids to pull sample_name from
sample_pool <- sample_pool %>%
  filter(!!sym(layer_1) %in% samples_to_select, n_samples >= 2) %>%
  select(!!sym(layer_1), !!sym(layer_2))

# Select sample_name to assess further
samples_to_select <- meta_data %>% 
  semi_join(sample_pool, by = c(layer_1, layer_2)) %>%
  pull(sample_name)

# Partition meta_data to contain viable samples
meta_data <- meta_data %>%
  filter(sample_name %in% samples_to_select)

# Partition expression matrix for samples in question
plot_data <- patient_data %>% 
  filter(patient_id != "control" & rownames(patient_data) %in% samples_to_select) %>%
              arrange(visit_id) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

de_genes <- run_group_by_split_DE(
     expr_mat = plot_data,
     meta = meta_data,
     split_var = layer_1,
     group_var = layer_2
 )

```

#Differential Gene Expression and Enrichment, Timepoint vs. All Cohorts
```{r gen_gene_enrich_timepoint_cohort_all, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

layer_1 <- "cohort_all"
layer_2 <- "visit_id"

meta_data <- patient_data %>% filter(patient_id != "control") %>%
              mutate(cohort_all = "all") %>%
              arrange(!!sym(layer_1)) %>% 
              select(c(!!sym(layer_1), !!sym(layer_2)))
meta_data$sample_name <- rownames(meta_data)

# ID the number of samples per cohort visit_id
sample_pool <- meta_data %>% 
  group_by(!!sym(layer_1), !!sym(layer_2)) %>% 
  tally(name = "n_samples")

# ID the visit_id with >=2 samples per visit_id and cohort
samples_to_select <- sample_pool %>%
  filter(n_samples >= 2) %>%
  group_by(!!sym(layer_1)) %>%
  tally(name = "n_groups") %>%
  filter(n_groups >= 2) %>%
  pull(!!sym(layer_1))

# ID the cohort and visit_ids to pull sample_name from
sample_pool <- sample_pool %>%
  filter(!!sym(layer_1) %in% samples_to_select, n_samples >= 2) %>%
  select(!!sym(layer_1), !!sym(layer_2))

# Select sample_name to assess further
samples_to_select <- meta_data %>% 
  semi_join(sample_pool, by = c(layer_1, layer_2)) %>%
  pull(sample_name)

# Partition meta_data to contain viable samples
meta_data <- meta_data %>%
  filter(sample_name %in% samples_to_select)

# Partition expression matrix for samples in question
plot_data <- patient_data %>% 
  filter(patient_id != "control" & rownames(patient_data) %in% samples_to_select) %>%
              arrange(visit_id) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

de_genes <- run_group_by_split_DE(
     expr_mat = plot_data,
     meta = meta_data,
     split_var = layer_1,
     group_var = layer_2
 )
```

#Differential Gene Expression and Enrichment, Timepoint vs. Cohorts 2-3 (all)
```{r gen_gene_enrich_timepoint_cohort_custom, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}

# First, run for all cohorts combined
layer_1 <- "cohort_all"
layer_2 <- "visit_id"

meta_data <- patient_data %>% filter(patient_id != "control" & cohort != "1") %>%
              mutate(cohort_all = "all") %>%
              arrange(!!sym(layer_1)) %>% 
              select(c(!!sym(layer_1), !!sym(layer_2)))
meta_data$sample_name <- rownames(meta_data)

# ID the number of samples per cohort visit_id
sample_pool <- meta_data %>% 
  group_by(!!sym(layer_1), !!sym(layer_2)) %>% 
  tally(name = "n_samples")

# ID the visit_id with >=2 samples per visit_id and cohort
samples_to_select <- sample_pool %>%
  filter(n_samples >= 2) %>%
  group_by(!!sym(layer_1)) %>%
  tally(name = "n_groups") %>%
  filter(n_groups >= 2) %>%
  pull(!!sym(layer_1))

# ID the cohort and visit_ids to pull sample_name from
sample_pool <- sample_pool %>%
  filter(!!sym(layer_1) %in% samples_to_select, n_samples >= 2) %>%
  select(!!sym(layer_1), !!sym(layer_2))

# Select sample_name to assess further
samples_to_select <- meta_data %>% 
  semi_join(sample_pool, by = c(layer_1, layer_2)) %>%
  pull(sample_name)

# Partition meta_data to contain viable samples
meta_data <- meta_data %>%
  filter(sample_name %in% samples_to_select)

# Partition expression matrix for samples in question
plot_data <- patient_data %>% 
  filter(patient_id != "control" & rownames(patient_data) %in% samples_to_select) %>%
              arrange(visit_id) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

de_genes <- run_group_by_split_DE(
     expr_mat = plot_data,
     meta = meta_data,
     split_var = layer_1,
     group_var = layer_2,
     output_path = paste0(working_dir, "cohort2-3/")
 )

# Run for cohort by cohort

layer_1 <- "cohort"
layer_2 <- "visit_id"

meta_data <- patient_data %>% filter(patient_id != "control" & cohort != "1") %>%
              arrange(!!sym(layer_1)) %>% 
              select(c(!!sym(layer_1), !!sym(layer_2)))
meta_data$sample_name <- rownames(meta_data)

# ID the number of samples per cohort visit_id
sample_pool <- meta_data %>% 
  group_by(!!sym(layer_1), !!sym(layer_2)) %>% 
  tally(name = "n_samples")

# ID the visit_id with >=2 samples per visit_id and cohort
samples_to_select <- sample_pool %>%
  filter(n_samples >= 2) %>%
  group_by(!!sym(layer_1)) %>%
  tally(name = "n_groups") %>%
  filter(n_groups >= 2) %>%
  pull(!!sym(layer_1))

# ID the cohort and visit_ids to pull sample_name from
sample_pool <- sample_pool %>%
  filter(!!sym(layer_1) %in% samples_to_select, n_samples >= 2) %>%
  select(!!sym(layer_1), !!sym(layer_2))

# Select sample_name to assess further
samples_to_select <- meta_data %>% 
  semi_join(sample_pool, by = c(layer_1, layer_2)) %>%
  pull(sample_name)

# Partition meta_data to contain viable samples
meta_data <- meta_data %>%
  filter(sample_name %in% samples_to_select)

# Partition expression matrix for samples in question
plot_data <- patient_data %>% 
  filter(patient_id != "control" & rownames(patient_data) %in% samples_to_select) %>%
              arrange(visit_id) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

de_genes <- run_group_by_split_DE(
     expr_mat = plot_data,
     meta = meta_data,
     split_var = layer_1,
     group_var = layer_2,
     output_path = paste0(working_dir, "cohort2-3/")
 )
```

# PCA
```{r run_pca, include = TRUE, echo = FALSE, fig.width = 5, fig.height = 5}

meta_data <- patient_data %>% 
  mutate(sample_name = rownames(patient_data)) %>%
  filter(patient_id != "control") %>%
  select(c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id")) %>%
  arrange(sample_name)

plot_data <- patient_data %>% 
  mutate(sample_name = rownames(patient_data)) %>%
  filter(patient_id != "control") %>%
  arrange(sample_name) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

pca_data <- run_pca(exp_matrix = plot_data,
        meta_data = meta_data,
        output_path = paste0(working_dir, "pca"))

gene_names <- c(pca_data$rotation[,"PC1"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC1"] %>% sort(decreasing = FALSE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC2"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC2"] %>% sort(decreasing = FALSE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC3"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC3"] %>% sort(decreasing = FALSE) %>% head(20) %>% names()
                ) %>% unique()

# Top 3 loadings
pca_loadings <- tibble(
  gene = gene_names,
  loading_pc1 = pca_data$rotation[gene_names, "PC1"],
  loading_pc2 = pca_data$rotation[gene_names, "PC2"],
  loading_pc3 = pca_data$rotation[gene_names, "PC3"]
  ) %>% arrange(desc(loading_pc1))

write.csv(x = pca_loadings,
          file = paste0(working_dir, "pca/pca_loadings_1-3.csv"))

write.csv(x = pca_data$rotation,
          file = paste0(working_dir, "pca/pca_loadings_all.csv"))

```

# PCA, Cohort 2 and 3 only
```{r run_pca_custom, include = TRUE, echo = FALSE, fig.width = 5, fig.height = 5}

meta_data <- patient_data %>% 
  mutate(sample_name = rownames(patient_data)) %>%
  filter(patient_id != "control" & cohort != "1") %>%
  select(c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id")) %>%
  arrange(sample_name)

plot_data <- patient_data %>% 
  mutate(sample_name = rownames(patient_data)) %>%
  filter(patient_id != "control" & cohort != "1") %>%
  arrange(sample_name) %>% 
              select(!c("sample_name", "cohort", "patient_id",
                        "visit_id", "order", "id", "cartridge", 
                        "batch", "timepoint", "set", "hours", "donor_id"))
plot_data <- plot_data %>% select(!c(colnames(plot_data)[grepl("^NEG_", colnames(plot_data))], 
                        colnames(plot_data)[grepl("^POS_", colnames(plot_data))]))

dir.create(paste0(working_dir, "cohort2-3/pca"), recursive = TRUE)
           
pca_data <- run_pca(exp_matrix = plot_data,
        meta_data = meta_data,
        output_path = paste0(working_dir, "cohort2-3/pca"))

gene_names <- c(pca_data$rotation[,"PC1"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC1"] %>% sort(decreasing = FALSE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC2"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC2"] %>% sort(decreasing = FALSE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC3"] %>% sort(decreasing = TRUE) %>% head(20) %>% names(),
                pca_data$rotation[,"PC3"] %>% sort(decreasing = FALSE) %>% head(20) %>% names()
                ) %>% unique()
# Top 3 loadings
pca_loadings <- tibble(
  gene = gene_names,
  loading_pc1 = pca_data$rotation[gene_names, "PC1"],
  loading_pc2 = pca_data$rotation[gene_names, "PC2"],
  loading_pc3 = pca_data$rotation[gene_names, "PC3"]
  ) %>% arrange(desc(loading_pc1))

write.csv(x = pca_loadings,
          file = paste0(working_dir, "cohort2-3/pca/pca_loadings_1-3.csv"))

write.csv(x = pca_data$rotation,
          file = paste0(working_dir, "cohort2-3/pca/pca_loadings_all.csv"))

```

# Cleanup Files
```{r run_cleanup, include = TRUE, echo = FALSE, fig.width = 5, fig.height = 5}

cleanup_files <- function(working_dir, data_dir, new_dir, file_pattern){
  
  if(!dir.exists(paste0(working_dir, file.path(data_dir, new_dir)))){
    dir.create(paste0(working_dir, file.path(data_dir, new_dir)), recursive = TRUE)
  }
  
  files_to_move <- list.files(path = paste0(working_dir, data_dir),
                            pattern = file_pattern,
                            full.names = TRUE)
  file.rename(from = files_to_move,
            to = file.path(paste0(working_dir, file.path(data_dir, new_dir)), 
                           basename(files_to_move)))
}

# Clean up cell_profiler files
cleanup_files(working_dir, "cell_profiler", "raw", "*_raw_*")
cleanup_files(working_dir, "cell_profiler", "relataive", "*_relative_*")

# Clean up dotplot_by_visit files
cleanup_files(working_dir, "dotplot_by_visit", "signature_cohort", "dotplot_cohort*")
cleanup_files(working_dir, "dotplot_by_visit", "signature_custom_all", "dotplot_custom_*")
cleanup_files(working_dir, "dotplot_by_visit", "signature_hours", "dotplot_hours_*")
cleanup_files(working_dir, "dotplot_by_visit", "signature_nanostring_all", "dotplot_nanostring_*")

# Clean up gsea files
cleanup_files(working_dir, "gsea", "visit_vs_visit_all", "*_all.png")
cleanup_files(working_dir, "gsea/visit_vs_visit_all", "baseline", "*_baseline_all.png")
cleanup_files(working_dir, "gsea/visit_vs_visit_all", "heatmap", "msig_c7_overrepresented_genes_heatmap_*")

cleanup_files(working_dir, "gsea", "visit_vs_visit_cohort", "msig_c7_overrepresented_genes_combined_visit_id_*")
cleanup_files(working_dir, "gsea", "visit_vs_visit_cohort", "msig_c7_overrepresented_genes_visit_id_*")
cleanup_files(working_dir, "gsea/visit_vs_visit_cohort", "baseline", "*_visit_id_baseline_*")
cleanup_files(working_dir, "gsea/visit_vs_visit_cohort/baseline", "heatmap", "*_heatmap_*")
cleanup_files(working_dir, "gsea", "visit_vs_visit_cohort/heatmap", "msig_c7_overrepresented_genes_heatmap_visit_id_*")

cleanup_files(working_dir, "gsea", "cohort_vs_cohort", "msig_c7_overrepresented_genes_cohort_*")
cleanup_files(working_dir, "gsea", "cohort_vs_cohort", "msig_c7_overrepresented_genes_combined_cohort_*")
cleanup_files(working_dir, "gsea", "cohort_vs_cohort/heatmap", "msig_c7_overrepresented_genes_heatmap_cohort_*")

cleanup_files(working_dir, "gsea", "csv", "*.csv")

# Clean up heatmap_by_order files
cleanup_files(working_dir, "heatmap_by_order", "all", "heatmap_log2fc_allgenes_*")
cleanup_files(working_dir, "heatmap_by_order", "custom", "heatmap_log2fc_custom_*")
cleanup_files(working_dir, "heatmap_by_order", "nanostring", "heatmap_log2fc_nanostring_*")

# Clean up volcanoplot files
cleanup_files(working_dir, "volcanoplot", "cohort_vs_cohort", "volcano_log2_cohort_*")
cleanup_files(working_dir, "volcanoplot", "visit_vs_visit_all", "*_all.png")
cleanup_files(working_dir, "volcanoplot/visit_vs_visit_all", "baseline", "*_baseline_all.png")
cleanup_files(working_dir, "volcanoplot", "visit_vs_visit_by_cohort", "volcano_log2_visit_id_*")
cleanup_files(working_dir, "volcanoplot/visit_vs_visit_by_cohort", "baseline", "*_baseline_*")

```

# Testing
```{r testing, include = TRUE, echo = FALSE, fig.width = 10, fig.height = 10}
```